---
title: "GCVieweR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GCVieweR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
devtools::load_all()
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

**GCVieweR** is available through GitHub

``` r
# install.packages("devtools")
devtools::install_github("nvelden/GCVieweR")
```

## Cluster Visualization

### Single Gene Cluster

To visualize a single gene cluster, the minimum requirements are the start and stop positions of each gene, typically named "start" and "stop". In addition a group can be defined which is used for color coding the genes and is used by default as the categorical variable for the legend.

```{r echo=TRUE, results='asis'}
# Example data for a single Gene cluster
ophA_cluster <- subset(ophA_clusters, cluster == "ophA")

GC_chart(ophA_cluster, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         height = "200px"
         )
```

### Multiple Gene Clusters

For visualizing multiple gene clusters, an additional variable is needed which is used to define the clusters (as shown in the 'cluster' column below). 

```{r echo=TRUE, results='asis'}
# Example data for two Gene clusters
GC_chart(ophA_clusters, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         cluster = "cluster", 
         height = "300px"
         )
```


## Customization

### Labels

To each gene cluster one can add a title, gene labels, a cluster label and a footer as shown in the example below. 

```{r echo=TRUE, results='asis'}
GC_chart(ophA_clusters, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         cluster = "cluster", 
         height = "420px"
         ) %>%
  GC_title(
    title = c("Gene Cluster"), 
    subtitle = c("Omphalotus Olearius", "Dendrothele Bisporus")
    ) %>%
  GC_labels("name") %>%
  GC_clusterLabel(title = unique(ophA_clusters$cluster)) %>%
  GC_footer(
    title = c("Nr. of Genes: 7", "Nr. of Genes: 10"), 
    subtitle = c("Locus: 2522 - 21,484", "Locus 19,236 - 43,005")
    )
```

To adjust the positioning of labels, the x, y, and position variables can be used. Note that in this example, GC_labels() is employed twice: initially, it sets labels uniformly for all gene clusters, and then it specifically adjusts the fourth label (considering the index starts at 0) in cluster 2, to ensure there is no overlap.

```{r echo=TRUE, results='asis'}
GC_chart(ophA_clusters, start = "start", stop = "stop", group = "class", cluster = "cluster", height = "360px") %>%
  GC_title(c("Omphalotus Olearius", "Dendrothele Bisporus"), position = "left", x = 20) %>%
  GC_labels("name", y = 25) %>%
  GC_labels("name", cluster = 2, y = 25, itemStyle = list(list(index = 3, y = 50))) %>%
  GC_footer(title = c("Nr. of Genes: 7", "Nr. of Genes: 10"), subtitle = c("Locus: 2522 - 21,484", "Locus 19,236 - 43,005"), position = "center", y = -30)
```

To adjust the position or styling of specific labels one can use the cluster and itemStyle variables. With the cluster variable one can select a specific cluster by their name or number. Using itemStyle variable one can provide a list with the index number of a specific label and the styles that should be applied to it. Note that the index for itemStyle starts with 0.

```{r echo=TRUE, results='asis'}
GC_chart(ophA_clusters, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         cluster = "cluster", 
         height = "360px") %>%
  GC_labels("name", y = 30) %>%
  GC_labels("name", 
            cluster = 2, 
            y = 30,
            itemStyle = list(
              list(index = 3, y = 50, fill = "red"),
              list(index = 7, y = 50, fill = "red")
              )
            )
```

For more detailed examples on how to adjust styling, positioning, and settings for these elements, see the Reference documentation of each function.

### Colors

The color of the legend and genes can be controlled through the `colorScheme` and `customColors` variables.

#### colorScheme

Defines a predefined color scheme for the legend, such as "schemeCategory10" or "schemeAccent" from D3.js. To maintain consistency in color representation apply the same `colorScheme` to both the legend and the gene elements. 

```{r echo=TRUE, results='asis'}
GC_chart(ophA_cluster, start = "start", stop = "stop", group = "class", height = "100px") %>%
  GC_legend(legendOptions = list(colorScheme = "schemeAccent")) %>%
  GC_genes(colorScheme = "schemeAccent")
```

#### customColors

Defines a predefined set of colors, customColors allows for specifying exact colors for each group in the legend. 

```{r echo=TRUE, results='asis'}
custom_color_map =c("#F8766D","#00BA38","#619CFF","#F564E3","#FFC61E","#00BFC4")

GC_chart(ophA_cluster, start = "start", stop = "stop", group = "class", height = "100px") %>%
  GC_legend(legendOptions = list(customColors = custom_color_map)) %>% 
  GC_genes(customColors = custom_color_map)
```

### Coordinates

Gene coordinates can be displayed with the `GC_coordinates()` function. When overlapping coordinates occur, they are, by default, shifted to the top axis for clearer visualization. By default the `overlapPercentage` parameter is set to 2%.

```{r echo=TRUE, results='asis'}
dbophA_cluster <- subset(ophA_clusters, cluster == "dbophA")

GC_chart(dbophA_cluster, start = "start", stop = "stop", group = "class",height = "100px") %>%
  GC_coordinates(
  rotate = -45,
  yPositionTop = 55,
  yPositionBottom = 45,
  overlapPercentage = 2
  ) %>%
  GC_legend(FALSE)
```

The style of the ticks and text can be customized using the tickStyle and textStyle variables

```{r echo=TRUE, results='asis'}
GC_chart(dbophA_cluster, start = "start", stop = "stop", group = "class",height = "100px") %>%
  GC_coordinates(
   tickStyle = list(stroke = "grey", strokeWidth = 1, lineLength = 10),
   textStyle = list(fill = "gray", fontSize = "10px", fontFamily = "Arial", cursor = "default")
   ) %>%
   GC_legend(FALSE)
```

By default, the start and stop positions of each gene are indicated by coordinates. However, you can customize these by providing your own values for the top and bottom ticks, using the tickValuesTop and tickValuesBottom options.

```{r echo=TRUE, results='asis'}
GC_chart(dbophA_cluster, start = "start", stop = "stop", group = "class",height = "100px") %>%
  GC_coordinates(
    tickValuesTop = c(27371, 28937),
    tickValuesBottom = c(25288, 27217)                  
   ) %>%
   GC_legend(FALSE)
```

### Scale 

#### Automatic Scale Breaks

In the `GC_chart()` function, the parameters `scale_breaks`, `scale_break_threshold`, and `scale_break_padding` can be used to avoid issues with widely spaced genes within a cluster. Setting `scale_breaks` to TRUE introduces breaks in the genomic scale which will avoid long empty stretches. The `scale_break_threshold` defines the spacing percentage between genes needed to activate these breaks. `scale_break_padding` controls the padding around these breaks.

In the given example, a scale break will be introduced in cases where the intergenic region exceeds 20% of the total range of the gene cluster. Additionally, a padding of 1% will be added to both the left and right sides of this break.

```{r echo=TRUE, results='asis'}
# With scale breaks
GC_chart(ophA_cluster, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         height = "100px", 
         scale_breaks = TRUE,
         scale_break_threshold = 20,
         scale_break_padding = 1) %>%
  GC_legend(FALSE)
```

#### Manual Scale Adjustments

By default the minimum and maximum of the genes start and stop positions are used to position the genes along the sequence. With the use of the `GC_scale()` function, these values can be adjusted. Furthermore, this function allows for the introduction of breaks within the sequence.

In the provided example, adjustments are made to the start and stop range to incorporate additional padding on both the left and right sides of the gene cluster. Additionally, a scale break is introduced to remove the extensive intergenic region that exists between gene 2 and gene 3.

```{r echo=TRUE, results='asis'}
GC_chart(
  ophA_cluster, 
  start = "start", 
  stop = "stop", 
  group = "class", 
  height = "100px") %>%
  GC_coordinates() %>%
  GC_scale(
    start = 1522, 
    stop = 22484,
    breaks = 
      list(
        list(start = 5486, stop = 9336)
      )
    ) %>%
  GC_legend(FALSE)
```

### Sequence

The sequence line in the `GC_sequence()` function is initially presented as a solid black line, but can be customized with the `sequenceStyle` option. Additionally, the `markerStyle` option allows adjusting the style of the break markers.

```{r echo=TRUE, results='asis'}
GC_chart(
  ophA_clusters, 
  start = "start", 
  stop = "stop", 
  group = "class",
  scale_breaks = TRUE,
  cluster = "cluster",
  height = "150px") %>%
  GC_sequence(
    show = TRUE,
    cluster = NULL,
    y = 50,
    sequenceStyle = list(
      stroke = "grey",
      strokeWidth = 1
      # Any other CSS style
    ),
    markerStyle = list(
      stroke = "grey",
      strokeWidth = 1,
      gap = 3,
      tiltAmount = 5
      # Any other CSS style
    )
  ) %>%
  GC_legend(FALSE)
```

### Scalebar

A scalebar can be added to the gene cluster chart using GC_scaleBar(). This function, by default, displays a scale bar of 1000 base pairs (bp) labeled as "1kb". 

```{r echo=TRUE, results='asis'}
GC_chart(
  ophA_clusters, 
  start = "start", 
  stop = "stop", 
  group = "class",
  cluster = "cluster",
  height = "150px") %>%
  GC_scaleBar(
    title = "1kb",
    scaleBarUnit = 1000
    ) %>%
  GC_legend(FALSE)
```

The scalebar's position and label can be customized using the `x`, `y`, and `labelPosition` parameters. CSS styles can be added to modify the labels. The appearance of the scalebar line and ticks is adjustable through the `scaleBarLine` and `scaleBarTick` options.

```{r echo=TRUE, results='asis'}
GC_chart(
  ophA_clusters, 
  start = "start", 
  stop = "stop", 
  group = "class",
  cluster = "cluster",
  height = "150px"
) %>%
  GC_scaleBar(
    title = "1kb",
    scaleBarUnit = 1000,
    cluster = NULL,
    x = 0,
    y = 10,
    labelStyle = list(
      labelPosition = "left",
      fontSize = "10px",
      fontFamily = "sans-serif",
      fill = "black",    
      cursor = "default" 
      # Any other CSS style for the label
    ),
    textPadding = 2,
    scaleBarLineStyle = list(
      stroke = "black",
      strokeWidth = 1
      # Any other CSS style for the line
    ),
    scaleBarTickStyle = list(
      stroke = "black",
      strokeWidth = 1   
      # Any other CSS style for the tick
    )
  )
```

### Tooltip

By default, the tooltip in the gene chart displays the start and stop positions of a gene when hovering over it. These positions are dynamically replaced by the actual start and stop data values for each gene. You can customize the text of the tooltip using the `formatter` variable. In the example below, the tooltip is customized to not only include the gene's start and stop positions but also its name. The gene name is pulled from the 'name' column in the data set. 

```{r echo=TRUE, results='asis'}
GC_chart(
  ophA_cluster, 
  start = "start", 
  stop = "stop", 
  group = "class", 
  height = "150px") %>%
  GC_tooltip(
    formatter = "<b>Gene: </b>{name}<br><b>Start:</b> {start}<br><b>Stop:</b> {stop}",
  ) %>%
  GC_legend(FALSE)
```

### Grid

The `GC_grid()` function can be used to modify the grid of all or a specific cluster. In the below example the height and margins are adjusted for the second gene cluster.

```{r echo=TRUE, results='asis'}
# With scale breaks
GC_chart(ophA_clusters, 
         start = "start", 
         stop = "stop", 
         group = "class", 
         cluster = "cluster",
         height = "200px" # Total height 
                          # since there are 2 clusters in the data 
                          # the height will be 200px each
         ) %>% 
  GC_grid(
    cluster = 2, # Set to NULL to adjust all clusters
    height = "80px",
    margin = list(left = "5px", right = "5px", top = "0px", bottom = "0px")
  ) %>% 
  GC_legend(FALSE)
```
