---
title: "Loading Gene Clusters From Genbank Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Gene Clusters From Genbank Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
library(genviewer)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Intro

This tutorial demonstrates creating gene cluster visualizations from GeneBank files using `geneviewer`. You can obtain GeneBank files containing genomic data from a variety of platforms such as [NCBI](https://www.ncbi.nlm.nih.gov/genbank/), [Ensembl](https://www.ensembl.org/), [UCSC](https://genome.ucsc.edu/), [antiSMASH](https://antismash.secondarymetabolites.org/#!/start) and [MIBiG](https://mibig.secondarymetabolites.org/). While `geneviewer` is designed to handle GeneBank files from these sources (please report any compatibility issues on [Github](https://github.com/nvelden/geneviewer/issues)), this tutorial specifically aims to replicate a gene cluster visualization found on the [MIBiG](https://mibig.secondarymetabolites.org/) website. 

### Loading GenBank file

The GenBank files from [MIBiG](https://mibig.secondarymetabolites.org/) include additional details, enabling us to color-code genes by function, which we will use to reproduce the visualizations they show on the website. Begin by downloading the GeneBank file from the [MIBiG](https://mibig.secondarymetabolites.org/) website for gene cluster [BGC0000001](https://mibig.secondarymetabolites.org/repository/BGC0000001/index.html#r1c1). Simply click the "Download GenBank file" link and save the file.

We can load the file into R using the `read_gbk` function from `geneviewer`. This will create a list containing all the sections and features found in the GenBank file. 

```{r eval=FALSE, results='hide'}
library(geneviewer)
# change the path to where you have saved the file
file_path <- "~/path/to/file/BGC0000001.gbk"
gbk <- read_gbk(file_path)
View(gbk) # Inspect the list in Rstudio
```

```{r echo = FALSE, results='asis'}
library(geneviewer)
# change the path to where you have saved the file
file_path <- "~/Documents/2023 github/geneviewer/genbank_files/BGC0000001.gbk"
gbk <- read_gbk(file_path)
```

We can inspect the .gbk using the View pane in Rstudio (use `View(gbk)`) or by simply opening the .gbk file in a text editor. For the visualization we are interested in the start and end position of each gene as well as their identifiers and functions. We can find these under FEATURES > CDS. Lets print the first CDS to see what information it contains.   

```{r echo=TRUE, results='hide'}
CDS_1 <- gbk[["BGC0000001"]][["FEATURES"]][["CDS"]][[1]] 
df <- data.frame(values = CDS_1)
df
```

```{r echo = FALSE, results='asis'}
CDS_1 <- gbk[["BGC0000001"]][["FEATURES"]][["CDS"]][[1]] 
df <- data.frame(values = CDS_1)
df["translation", "values"] <- substr(df["translation", "values"], 1, 40)
rownames(df) <- names(CDS_1)
kable(df)
```

We can bind the information of each CDS into a data.frame using the `gbk_features_to_df` function. In the below code we define the feature and the keys we are interested in and bind it into a data.frame. We set process_region to `TRUE` which will extract the strand, start and end from the region key.

```{r echo=TRUE, results='hide'}
cds_df <- gbk_features_to_df(
  gbk, 
  feature = "CDS",
  keys = c("region", "gene", "protein_id", "gene_kind", "product"), 
  process_region = TRUE
  )
```

```{r echo = FALSE, results='asis'}
cds_df$region <- NULL
kable(head(cds_df))
```

### Cluster Visualization

We can see that there are some NA values in the gene_kind column so we will replace those with "other". After we have done that we are ready to make the cluster visualization.

```{r echo = FALSE, results='asis'}
# replace NA with "other" for gene_kind
cds_df$gene_kind[is.na(cds_df$gene_kind)] <- "other"

# Make the cluster chart
chart <- GC_chart(
  cds_df,
  start = "start",
  end = "end",
  group = "gene_kind",
  strand = "strand",
  height = "130px"
)
chart
```

To make the chart more similar to the one we find on the [MIBiG](https://mibig.secondarymetabolites.org/) website we can add a title, alter the coloring and appearance of the genes and add a scale. In addition, we alter the tooltip to display the gene, product and region on hoover. 

```{r echo = FALSE, results='asis'}
colors <- list(
  "biosynthetic" = "#810E15", 
  "biosynthetic-additional" = "#F16D75", 
  "transport" = "#6495ED",
  "regulatory" = "#2E8B57", 
  "other" = "#808080"
)

chart %>%
  GC_genes(
    arrowheadWidth = 10,
    arrowheadHeight = 15,
    arrowHeight = 15
  ) %>%
  GC_title(
    "BGC0000001: abyssomicin C biosynthetic gene cluster
    from <i>Verrucosispora maris</i> AB-18-032",
    align = "left",
    titleFont = list(
      fontSize = "12px"
    )
    ) %>%
  GC_scale() %>%
  GC_color(customColors = colors) %>%
  GC_legend(order = names(colors)) %>%
  GC_tooltip(
    formatter = "
    <b>Gene:</b> {gene}<br>
    <b>Product:</b> {product}<br>
    <b>Region:</b>{start} - {end}
    "
  )
```

See the example section to see how to further customize cluster charts.
